---
title: "Data Exercise - Vaccine Adverse Event Analysis"
author: "Asmith Joseph"
date: 1/30/2025
format: html
---

## Welcome to the Data Exercise!



# **Introduction: Vaccine Adverse Event Analysis**
#### Vaccination is one of the most effective public health measures to prevent infectious diseases. However, like any medical intervention, vaccines can sometimes lead to adverse events, ranging from mild side effects to severe complications. This analysis explores data from the Vaccine Adverse Event Reporting System (VAERS) to understand trends, patterns, and factors influencing adverse vaccine reactions.


# **Data Import & Loading**
```{r}
# **2. Data Import & Loading**! 
# Uploading the data. We are using the the Vaccine Adverse Event Reporting System (VAERS) dataset
# Load required library
library(readr)

# Define the correct file path
file_path <- "C:/Users/ajose35/Desktop/Ecole/MADA-course/AsmithJoseph-MADA-portfolio/data-exercise/VaccineData.csv"

# Load the data
VaccineData <- read_csv(file_path)

# Display first few rows
head(VaccineData)

# Optional: View the dataset in RStudio
View(VaccineData)


```


# **Initial Data Exploration**
```{r}
#Looking at the structure of the data to verify where there issues and start the cleaning process

# Viewieng the dataset
View(VaccineData)



str(VaccineData)  
```



# **Data Cleaning & Preprocessing**
```{r}
# There are several issues, such as dates need conversion to proper format, missing values and so on. So I am starting the cleaning process. 
# Printing column names to check if they match what we are trying to mutate
colnames(VaccineData)


# Loading some of the  necessary libraries
library(dplyr)
library(lubridate)

# Converting character dates to proper Date format
VaccineData <- VaccineData %>%
  mutate(
    RECVDATE = mdy(RECVDATE),   # Convert from MM/DD/YYYY to Date
    RPT_DATE = mdy(RPT_DATE),
    VAX_DATE = mdy(VAX_DATE),
    ONSET_DATE = mdy(ONSET_DATE),
    DATEDIED = mdy(DATEDIED),
    TODAYS_DATE = mdy(TODAYS_DATE)
  )

# Checking structure after conversion
str(VaccineData)

```



```{r}

# Now I am checking for missing values 
# Count missing values per column
missing_counts <- colSums(is.na(VaccineData))

# Show only columns with missing values
missing_counts[missing_counts > 0]

```


```{r}
# Defining the columns I originally wanted to drop
columns_to_drop <- c("ER_VISIT", "BIRTH_DEFECT", "OFC_VISIT", "ER_ED_VISIT", "DATEDIED", "LAB_DATA")

# Keeping only the columns that exist in VaccineData
columns_to_drop <- columns_to_drop[columns_to_drop %in% colnames(VaccineData)]

# Dropping only the valid columns
VaccineData <- VaccineData %>% select(-all_of(columns_to_drop))

# Verifying the new dataset structure
colnames(VaccineData)


```



```{r}
# Dropping rows where AGE_YRS or STATE is missing (since these are critical for analysis)
VaccineData <- VaccineData %>%
  filter(!is.na(AGE_YRS) & !is.na(STATE))



# Replacing missing AGE_YRS with the median value
VaccineData$AGE_YRS[is.na(VaccineData$AGE_YRS)] <- median(VaccineData$AGE_YRS, na.rm = TRUE)


# Replacing missing STATE values with "Unknown"
VaccineData$STATE[is.na(VaccineData$STATE)] <- "Unknown"



colSums(is.na(VaccineData))


```



```{r}
# I am dropping columns with too many missing values. specifically over 70% missing.
# Defining the  threshold for missing values (70%)

"HOSPITAL" %in% colnames(VaccineData) 


threshold <- 0.7 * nrow(VaccineData)

# Identifying columns with excessive missing values (excluding "HOSPITAL")
cols_to_drop <- names(VaccineData)[colSums(is.na(VaccineData)) > threshold]

# Remove "HOSPITAL" from the drop list if it appears
cols_to_drop <- setdiff(cols_to_drop, "HOSPITAL")  

# Dropping columns ONLY if there are any that meet the threshold
if (length(cols_to_drop) > 0) {
  VaccineData <- VaccineData %>% select(-all_of(cols_to_drop))
}

# Printing remaining column names to confirm
colnames(VaccineData)


```



```{r}
# Print the dropped columns
print(cols_to_drop)

# Count how many columns were removed
length(cols_to_drop)



str(VaccineData)


```






```{r}
# Now that I have done with the cleaning process, which included: missing values, date & categorical variables formatted corrcetly, no extreme outliers or duplicates records. I will now proceed to explore the data by focusing on patterns, trends, and distributions using summary statistics and visualizations.


# Summary of numeric columns
summary(select(VaccineData, AGE_YRS, NUMDAYS, FORM_VERS))  #


```



# **Visualization of Age Distribution of Patients**
```{r}
#Visualization of Age Distribution of Patients
library(ggplot2)

ggplot(VaccineData, aes(x = AGE_YRS)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Age Distribution of Patients",
       x = "Age (years)",
       y = "Count") +
  theme_minimal()

```



# **Exploratory Data Analysis**
```{r}
# I want to explore the data more. First I am doing an age group analysis to find out whether or not certain age groups more likely to report adverse effects.
# Load required libraries
library(dplyr)
library(ggplot2)

# Creating age groups
VaccineData <- VaccineData %>%
  mutate(AgeGroup = case_when(
    AGE_YRS < 18 ~ "Child (0-17)",
    AGE_YRS >= 18 & AGE_YRS <= 40 ~ "Young Adult (18-40)",
    AGE_YRS > 40 & AGE_YRS <= 60 ~ "Middle-aged (41-60)",
    AGE_YRS > 60 ~ "Elderly (60+)"
  ))

# Counting reports per age group
AgeDistribution <- VaccineData %>%
  group_by(AgeGroup) %>%
  summarise(Count = n())



```



# **Visualization**

```{r}
# Ploting the distribution
ggplot(AgeDistribution, aes(x = AgeGroup, y = Count, fill = AgeGroup)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Adverse Events by Age Group",
       x = "Age Group",
       y = "Number of Reports") +
  theme_minimal()
```

```{r}
# A second Exploratory Data Analysis to identify whether men or women more likely to report adverse effects.
# Count reports by sex
SexDistribution <- VaccineData %>%
  group_by(SEX) %>%
  summarise(Count = n())

# Plot the distribution
ggplot(SexDistribution, aes(x = SEX, y = Count, fill = SEX)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Adverse Events by Sex",
       x = "Sex",
       y = "Number of Reports") +
  theme_minimal()

```



```{r}
print(SexDistribution)
```




```{r}
# Simplify CUR_ILL categories
VaccineData <- VaccineData %>%
  mutate(CUR_ILL_Grouped = case_when(
    grepl("asthma|respiratory|bronchitis", CUR_ILL, ignore.case = TRUE) ~ "Respiratory",
    grepl("diabetes", CUR_ILL, ignore.case = TRUE) ~ "Diabetes",
    grepl("hypertension|high blood pressure", CUR_ILL, ignore.case = TRUE) ~ "Hypertension",
    grepl("cancer", CUR_ILL, ignore.case = TRUE) ~ "Cancer",
    grepl("allergy|allergies", CUR_ILL, ignore.case = TRUE) ~ "Allergies",
    grepl("arthritis", CUR_ILL, ignore.case = TRUE) ~ "Arthritis",
    is.na(CUR_ILL) ~ "No Known Condition",
    TRUE ~ "Other"
  ))






# Count hospitalization cases by grouped pre-existing conditions
IllnessImpact <- VaccineData %>%
  group_by(CUR_ILL_Grouped, HOSPITAL) %>%
  summarise(Count = n())

# Print the table
print(IllnessImpact)


```





```{r}
# Plot impact of pre-existing conditions on hospitalization
ggplot(IllnessImpact, aes(x = CUR_ILL_Grouped, y = Count, fill = HOSPITAL)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Impact of Pre-existing Illness on Hospitalization",
       x = "Pre-existing Condition",
       y = "Number of Cases",
       fill = "Hospitalized") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




```{r}
# Create a contingency table
sex_hospital_table <- table(VaccineData$SEX, VaccineData$HOSPITAL)

# Run chi-square test
chisq.test(sex_hospital_table)

```




```{r}
# Count hospitalizations by sex
SexHospitalization <- VaccineData %>%
  group_by(SEX, HOSPITAL) %>%
  summarise(Count = n(), .groups = "drop")

# Plot hospitalization rates by sex
ggplot(SexHospitalization, aes(x = SEX, y = Count, fill = HOSPITAL)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Hospitalization by Sex",
       x = "Sex",
       y = "Number of Cases",
       fill = "Hospitalized") +
  theme_minimal()

```







```{r}
#On this last section we want to identify "What factors influence the likelihood of recovery from a vaccine adverse event?"

summary(select(VaccineData, AGE_YRS, NUMDAYS, FORM_VERS))

table(VaccineData$SEX)

table(VaccineData$RECOVD, useNA = "ifany")  # Shows how many "Y", "N", and missing values exist

table(VaccineData$SEX, VaccineData$RECOVD)

```






```{r}
#Plotting Histogram of Age Distribution
library(ggplot2)

ggplot(VaccineData, aes(x = AGE_YRS)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Age Distribution of Patients", x = "Age (years)", y = "Count") +
  theme_minimal()

```






```{r}
#Plotting a Bar Chart of  Recovery Status by Gender
ggplot(VaccineData, aes(x = SEX, fill = RECOVD)) +
  geom_bar(position = "dodge") +
  labs(title = "Recovery Status by Gender", x = "Sex", y = "Count") +
  theme_minimal()

```



```{r}
# Plotting a Boxplot of Age vs. Recovery
ggplot(VaccineData, aes(x = RECOVD, y = AGE_YRS, fill = RECOVD)) +
  geom_boxplot() +
  labs(title = "Age Distribution by Recovery Status", x = "Recovery", y = "Age (Years)") +
  theme_minimal()

```







```{r}
#Plotting a Boxplot of  Symptom Duration vs. Recovery
ggplot(VaccineData, aes(x = RECOVD, y = NUMDAYS, fill = RECOVD)) +
  geom_boxplot() +
  labs(title = "Symptom Duration by Recovery Status", x = "Recovery", y = "Days Until Recovery") +
  theme_minimal()

```




```{r}
# Selecting relevant variables for the model 
model_data <- VaccineData %>%
  select(RECOVD, AGE_YRS, SEX, NUMDAYS) %>%
  filter(!is.na(RECOVD) & !is.na(AGE_YRS) & !is.na(SEX) & !is.na(NUMDAYS)) %>%
  filter(RECOVD %in% c("Y", "N"))  # Remove "U" (unknown values)

# Converting categorical variables to factors
model_data <- model_data %>%
  mutate(RECOVD = as.factor(RECOVD),
         SEX = as.factor(SEX))

# Checking dataset summary
str(model_data)
table(model_data$RECOVD)  # Ensure we have two classes (Y/N)

```



# **Predictive Modeling: Logistic Regression for Recovery**
```{r}
# Fitting a Logistic Regression Model to identify "What factors influence the likelihood of recovery from a vaccine adverse event?"
#Older individuals are less likely to recover.
#Males are slightly more likely to recover than females.
#Longer symptom duration reduces recovery probability


recovery_model <- glm(RECOVD ~ AGE_YRS + SEX + NUMDAYS,
                      data = model_data,
                      family = "binomial")

# Print Model Summary
summary(recovery_model)

```


# **Conclusion**
#### This analysis explored factors influencing recovery from vaccine adverse events using VAERS data and logistic regression.                             

#### Key findings:                                                                                                                                             1) Age negatively impacts recovery (older individuals recover less, p < 2e-16).                                                                            2) Males recover slightly more than females (p = 1.88e-10).                                                                                                3) Longer symptom duration reduces recovery probability (p = 1.44e-14).


















